---
title: "Intrinsic Potential for Chinook Salmon in the Klamath Basin"
author: "Maddee Wiggins (FlowWest)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
library(igraph)
library(rivermile)

```

## Objective

Calculate the intrinsic potential for Chinook salmon in the Klamath basin rivers and streams

### Flowlines

Source: - NHDplus geodatabase downloaded from: <https://www.sciencebase.gov/catalog/item/5d30c29ae4b01d82ce84aa5e> - NHDplus user guide: <https://pubs.usgs.gov/of/2019/1096/ofr20191096.pdf>

```{r}
flowlines <- rivermile::processed_nhd 
```

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines, color = "darkblue", alpha = 0.5) +
  theme_minimal()
```

```{r}
flowlines_ip_layers <- flowlines |> select(gnis_nm, gnis_id, slope, tot_da_sq_km = tt_d_s_)
```

### Slope

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers, aes(color = slope)) +
  theme_minimal()
```

### Total Drainage Area (sqkm)

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers, aes(color = tot_da_sq_km)) +
  theme_minimal()
```

### Estimated Discharge

```{r}
estimate_Q <- function(area_km2) {
  Q <- 0.028 * area_km2^0.89
  return(Q)
}

flowlines_ip_layers <- flowlines_ip_layers |> 
  mutate(q_mean_est = estimate_Q(tot_da_sq_km))

```

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers, aes(color = q_mean_est)) +
  theme_minimal()
```

### Bankfull Width

Source: Bieger model $2.76 * DA^.399$

```{r}

flowlines_ip_layers <- flowlines_ip_layers |> 
  mutate(bf_width_m = 2.76 * tot_da_sq_km^0.399)
```

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers, aes(color = bf_width_m)) +
  theme_minimal()
```

### Valley Bottom Width

could use this workflow from habistat: <https://github.com/FlowWest/habistat/blob/main/data-raw/scripts/terrain-attributes.md#valley-bottom-width-via-slope-cutoff-method>

### Valley Width Index

$V = (valley-floor-width)/(active-channel-width)$

### Suitability Criteria

source: <https://www.fs.usda.gov/pnw/pubs/journals/pnw_2005_agrawal001.pdf>

```{r}
chinook_suit <- tibble(
  variable = c(
    rep("gradient_percent", 3),
    rep("valley_constraint", 3),
    rep("discharge_m3s", 3)
  ),
  value = c(
    0, 3.7, 8,        
    1, 4, 8,        
    0, 100, 100          
  ),
  suitability = c(
    1.0, 0.0, 0.0,  
    0.8, 1.0, 1.0,  
    1, 1.0, 1.0        
  )
)


ggplot() + 
  geom_line(data = chinook_suit, aes(x = value, y = suitability)) +
  facet_wrap(~variable, scales = "free_x") +
  theme_minimal()

```

```{r}
knitr::knit_exit()
```

```{r}
nhd_flowline <- readRDS(here::here("data-raw", "shapes", "nhd_flowline_hr.rds"))

flowlines <- nhd_flowline |>
  st_set_geometry("shape") |> 
  st_intersection(klamath_hucs |> st_union())

names(flowlines)[names(flowlines) == "shape"] <- "geometry"
st_geometry(flowlines) <- "geometry"
```

```{r}
flow_net <- flowlines 
seed_comid <- 50000400214915
seed_row <- flow_net |>
  st_drop_geometry() |>
  filter(comid == seed_comid)

edges <- flow_net |>
  st_drop_geometry() |>
  select(comid, from_node, to_node) |> 
  filter(!is.na(to_node))

g <- graph_from_data_frame(
  d = edges |> select(from_node, to_node),
  directed = TRUE
)

comp <- components(g, mode = "weak")

node_membership <- tibble(
  node = names(comp$membership),
  component_id = as.integer(comp$membership)
)

flow_comp <- flow_net |>
  left_join(
    node_membership |> 
      rename(from_node = node) |> 
      mutate(from_node = as.numeric(from_node)),
    by = "from_node"
  )

seed_component <- flow_comp |>
  st_drop_geometry() |>
  filter(comid == seed_comid) |>
  pull(component_id)

flow_clean <- flow_comp |>
  filter(component_id == seed_component)

flow_filter <- flow_clean |> 
  filter(stream_orde >= 5) |> 
  filter(tot_da_sq_km >= 1)

# flowlines_clean <- flow_clean |> 
#   select(comid, gnis_name, lengthkm, stream_leve, stream_orde, tot_da_sq_km, slope, geometry) |> 
#   filter(gnis_name %in% rivermile::all_klamath_rivers_line$river)
```

```{r}
ggplot() +
  geom_sf(data = flow_filter) 


# library(leaflet)
# 
# leaflet() |>
#   addTiles() |>
#   addPolylines(
#     data = flow_filter |> st_transform(4326),
#     popup = ~as.character(gnis_name)
#   )
```
