---
title: "Intrinsic Potential for Chinook Salmon — Klamath Basin"
author: "Maddee Wiggins (FlowWest)"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(sf)
library(rivermile)
library(terra)
library(stars)
library(ggrepel)
library(classInt)
library(here)
```

## Objective

Estimate the intrinsic potential (IP) for fall Chinook salmon spawning across the Klamath Basin stream network. Following Bjorkstedt et al. (2005), IP is modeled at two scales:

- **IP_reach** — a reach-level geometric mean of three physical suitability scores (gradient, discharge, and valley confinement), structured as a limiting-factors analysis in which very low suitability on any single variable greatly reduces the overall score regardless of the other two.
- **IP-km** — a watershed-level metric equal to the length-weighted sum of IP_reach across all accessible reaches, expressing a watershed's expected equivalent length of continuous high-quality habitat.

All reaches are evaluated regardless of survey coverage, treating the absence of fish observations as a gap in knowledge rather than evidence of absence.

---

## Data Aggregation and Processing

### Flowlines

Stream network geometry and baseline reach attributes (slope, cumulative drainage area) are sourced from NHDPlus.

- NHDplus geodatabase: <https://www.sciencebase.gov/catalog/item/5d30c29ae4b01d82ce84aa5e>
- NHDplus user guide: <https://pubs.usgs.gov/of/2019/1096/ofr20191096.pdf>

```{r load-flowlines, include=FALSE}
flowlines_raw <- rivermile::processed_nhd

flowlines_ip <- flowlines_raw |>
  select(gnis_nm, gnis_id, nhdplusid = nhd_pl_, slope, tot_da_sq_km = tt_d_s_) |>
  mutate(slope = slope * 100)  # convert fraction to percent
```

```{r map-network, fig.cap="NHDPlus stream network within the Klamath Basin study area."}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union(), fill = NA, color = "grey50") +
  geom_sf(data = flowlines_raw, color = "steelblue", linewidth = 0.3, alpha = 0.6) +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(), panel.grid = element_blank()) +
  labs(title = "Klamath Basin Stream Network")
```

### Estimated Discharge

Mean annual discharge is estimated from cumulative upstream drainage area using a regional power-law regression:

$$Q = 0.028 \cdot A^{0.89}$$

where $A$ is total drainage area in km². Because Chinook suitability is 1.0 for any discharge ≥ 3 m³/s, a more precise discharge estimate is unlikely to alter IP classifications substantially for the larger named reaches.

```{r discharge, include=FALSE}
flowlines_ip <- flowlines_ip |>
  mutate(discharge_mean_est = 0.028 * tot_da_sq_km^0.89)
```

### Bankfull Width

Bankfull width is estimated with the Bieger et al. (2015) regional equation:

$$W_{bf} = 2.76 \cdot A^{0.399}$$

where $A$ is total drainage area (km²). Bankfull width enters the IP model indirectly as the denominator of the valley width index.

```{r bankfull-width, include=FALSE}
flowlines_ip <- flowlines_ip |>
  mutate(bf_width_m = 2.76 * tot_da_sq_km^0.399)
```

### Valley Bottom Width

Valley bottom width is derived from a 12-m DEM using the slope-cutoff method: raster cells with slope < 14.04° (tan⁻¹(0.25)) are classified as valley bottom, and mean width is measured via perpendicular transects along each reach flowline. The methodology follows the [South Fork Eel River Basin Geomorphic Classification](https://watermanagement.ucdavis.edu/download_file/view_inline/144) and the implementation in [habistat](https://github.com/FlowWest/habistat).

```{r valley-fns, include=FALSE}
valley_bottom <- function(catchment, flowline, dem) {
  dem |>
    st_crop(catchment) |>
    st_as_stars() |>
    starsExtra::slope() |>
    mutate(valley_bottom = if_else(
      slope < units::set_units(atan(0.25) * 180 / pi, "degrees"), 1, NA
    )) |>
    select(valley_bottom) |>
    st_as_sf(merge = TRUE) |>
    st_filter(st_zm(flowline))
}

valley_bottom_by_comid <- function(x, dem, catchment, flowlines) {
  c <- catchment |> mutate(nhdplusid = as.character(nhdplusid)) |>
    filter(nhdplusid == x) |> st_transform(st_crs(dem))
  f <- flowlines |> mutate(nhdplusid = as.character(nhdplusid)) |>
    filter(nhdplusid == x) |> st_transform(st_crs(dem))
  valley_bottom(c, f, dem)
}

perpendicular_transect <- function(pair, transect_length) {
  p1 <- st_coordinates(st_sfc(pair[[1]])); p2 <- st_coordinates(st_sfc(pair[[2]]))
  mx <- (p1[1,"X"] + p2[1,"X"]) / 2; my <- (p1[1,"Y"] + p2[1,"Y"]) / 2
  pb <- atan2(p2[1,"Y"] - p1[1,"Y"], p2[1,"X"] - p1[1,"X"]) + pi / 2
  r  <- transect_length / 2
  st_linestring(matrix(
    c(mx + r*cos(pb), my + r*sin(pb), mx - r*cos(pb), my - r*sin(pb)),
    ncol = 2, byrow = TRUE
  ))
}

perpendicular_transects <- function(ls, transect_length) {
  if (inherits(ls, "sf"))       { crs_use <- st_crs(ls); ls <- st_geometry(ls) }
  else if (inherits(ls, "sfc")) { crs_use <- st_crs(ls) }
  else if (inherits(ls, "sfg")) { crs_use <- NA; ls <- st_sfc(ls) }
  else stop("ls must be sf, sfc, or sfg")
  coords <- st_coordinates(st_zm(ls))[, 1:2, drop = FALSE]
  if (nrow(coords) < 2) stop("Need at least 2 coordinates to build transects")
  pairs <- lapply(
    seq_len(nrow(coords) - 1),
    function(i) list(st_point(coords[i, ]), st_point(coords[i + 1, ]))
  )
  st_sfc(lapply(pairs, perpendicular_transect, transect_length = transect_length), crs = crs_use)
}

valley_width_via_transects <- function(geom_vb, flowline, max_width) {
  transects <- perpendicular_transects(flowline, transect_length = max_width)
  if ("sfg" %in% class(geom_vb)) geom_vb <- st_sfc(geom_vb, crs = st_crs(transects))
  mean(st_length(st_intersection(transects, geom_vb)))
}
```

```{r load-dem-catchments, include=FALSE}
dem        <- stars::read_stars("../../klamathAgData/data-raw/DEM/dem_z12_utm10.tif")
catchments <- rivermile::nhd_catchments |>
  mutate(nhdplusid = as.character(nhdplusid)) |>
  st_transform(st_crs(dem))
```

Valley-width calculations are computationally intensive and have been pre-computed in batches. The chunk below is retained for reproducibility but is not evaluated during rendering.

```{r compute-valley-width, eval=FALSE, include=FALSE}
safe_vb_transect <- purrr::possibly(
  function(a, b) {
    out <- valley_width_via_transects(a, b, 1000)
    if (length(out) == 0 || is.nan(out)) NA_real_ else as.numeric(out)
  },
  otherwise = NA_real_
)

safe_vb_simple <- purrr::possibly(
  function(x) {
    out <- as.numeric(terra::width(terra::vect(x)))
    if (length(out) == 0 || is.nan(out)) NA_real_ else out
  },
  otherwise = NA_real_
)

vb_fun <- function(data) {
  data |>
    mutate(
      nhdplusid = as.character(nhdplusid),
      flowline  = map(
        geometry,
        ~ st_sfc(st_zm(.x), crs = st_crs(flowlines_ip)) |> st_transform(st_crs(dem))
      )
    ) |>
    st_drop_geometry() |>
    mutate(valley_bottom_geoms = map(
      nhdplusid,
      purrr::possibly(
        function(x) {
          vb <- valley_bottom_by_comid(x, dem, catchment = catchments, flowlines = flowlines_ip)
          st_set_crs(st_union(st_geometry(vb)), st_crs(dem))
        },
        otherwise = st_sfc(st_polygon(), crs = st_crs(dem))
      )
    )) |>
    mutate(
      vb_width_simple   = map_dbl(valley_bottom_geoms, safe_vb_simple),
      vb_width_transect = map2_dbl(valley_bottom_geoms, flowline, safe_vb_transect)
    ) |>
    select(nhdplusid, flowline, valley_bottom_geoms, vb_width_simple, vb_width_transect)
}

batches <- list(
  list(rows = 1:500,      file = "valley_bottom_width_1_500.rds"),
  list(rows = 501:1000,   file = "valley_bottom_width_501_1000.rds"),
  list(rows = 1001:1500,  file = "valley_bottom_width_1001_1500.rds"),
  list(rows = 1501:2000,  file = "valley_bottom_width_1501_2000.rds"),
  list(rows = 2001:3000,  file = "valley_bottom_width_2001_3000.rds"),
  list(rows = 3001:4000,  file = "valley_bottom_width_3001_4000.rds"),
  list(rows = 4001:4326,  file = "valley_bottom_width_4001_4236.rds")
)

for (b in batches) {
  saveRDS(
    vb_fun(flowlines_ip |> slice(b$rows)),
    here::here("data-raw", b$file)
  )
}
```

```{r load-valley-width, include=FALSE}
vb_files <- c(
  "valley_bottom_width_1_500.rds",    "valley_bottom_width_501_1000.rds",
  "valley_bottom_width_1001_1500.rds","valley_bottom_width_1501_2000.rds",
  "valley_bottom_width_2001_3000.rds","valley_bottom_width_3001_4000.rds",
  "valley_bottom_width_4001_4236.rds"
)

vb <- bind_rows(lapply(vb_files, function(f) read_rds(here::here("data-raw", f))))

flowlines_ip <- flowlines_ip |>
  mutate(nhdplusid = as.character(nhdplusid)) |>
  left_join(vb |> select(nhdplusid, vb_width_simple, vb_width_transect),
            by = "nhdplusid")
```

### Valley Width Index

The valley width index (VWI) is the ratio of valley bottom width to bankfull channel width:

$$VWI = \frac{W_{valley}}{W_{bf}}$$

VWI serves as a proxy for lateral confinement: values < 1 indicate a fully confined channel; values > 4 indicate an unconfined, laterally mobile channel where gravel recruitment and retention are favorable.

```{r valley-width-index, include=FALSE}
flowlines_ip <- flowlines_ip |>
  mutate(valley_width_index = vb_width_transect / bf_width_m)
```

```{r vwi-map, fig.cap="Valley Width Index across the Klamath Basin stream network. Lighter values indicate more unconfined reaches."}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union(), fill = NA, color = "grey50") +
  geom_sf(data = flowlines_ip |> filter(!is.na(valley_width_index)),
          aes(color = pmin(valley_width_index, 10)), linewidth = 0.5) +
  scale_color_viridis_c(name = "VWI\n(capped at 10)", option = "mako", direction = -1) +
  theme_minimal(base_size = 12) +
  theme(axis.title = element_blank(), panel.grid = element_blank())
```

```{r vwi-summary}
flowlines_ip |>
  st_drop_geometry() |>
  group_by(gnis_nm) |>
  summarise(
    Min    = round(min(valley_width_index, na.rm = TRUE), 1),
    Median = round(median(valley_width_index, na.rm = TRUE), 1),
    Mean   = round(mean(valley_width_index, na.rm = TRUE), 1),
    Max    = round(max(valley_width_index, na.rm = TRUE), 1),
    `% Missing` = round(mean(is.na(valley_width_index)) * 100, 0)
  ) |>
  knitr::kable(caption = "Valley Width Index summary by stream.")
```

---

## Suitability Criteria

Suitability scores are expressed on a continuous 0–1 scale (1 = fully suitable, 0 = unsuitable). Thresholds are drawn from Bjorkstedt et al. (2005), Burnett et al. (2007), Busch et al. (2013), Connor et al. (2015), and Beechie et al. (2006).

```{r suitability-curves, fig.cap="Suitability step functions for gradient (%), discharge (m³/s), and valley width index used in the IP model."}
suit_curves <- tribble(
  ~variable,                ~value, ~suitability,
  "Gradient (%)",              0.0,          1.0,
  "Gradient (%)",              2.0,          1.0,
  "Gradient (%)",              2.0,         0.45,
  "Gradient (%)",              4.0,         0.45,
  "Gradient (%)",              4.0,          0.0,
  "Gradient (%)",             10.0,          0.0,
  "Discharge (m³/s)",          0.0,          0.0,
  "Discharge (m³/s)",          1.0,          0.0,
  "Discharge (m³/s)",          1.0,         0.45,
  "Discharge (m³/s)",          3.0,         0.45,
  "Discharge (m³/s)",          3.0,          1.0,
  "Discharge (m³/s)",         20.0,          1.0,
  "Valley Width Index",        0.0,          0.0,
  "Valley Width Index",        1.0,          0.0,
  "Valley Width Index",        1.0,         0.45,
  "Valley Width Index",        4.0,         0.45,
  "Valley Width Index",        4.0,          1.0,
  "Valley Width Index",       12.0,          1.0
)

ggplot(suit_curves, aes(x = value, y = suitability)) +
  geom_step(linewidth = 1, color = "steelblue", direction = "hv") +
  facet_wrap(~variable, scales = "free_x") +
  scale_y_continuous(limits = c(-0.05, 1.1), breaks = c(0, 0.45, 1)) +
  labs(x = NULL, y = "Suitability (0–1)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

---

## Intrinsic Potential

### Method

Following Bjorkstedt et al. (2005), the reach-level IP score is the geometric mean of the three suitability scores:

$$IP_{reach} = \left(S_{gradient} \times S_{discharge} \times S_{confinement}\right)^{1/3}$$

The geometric mean gives the model the structure of a limiting-factors analysis: very low suitability on any single variable greatly reduces or eliminates the probability of suitable habitat, regardless of how favorable the other two variables are.

At the watershed scale, IP is summarized as IP-km — the length-weighted sum of reach-level scores:

$$IP\text{-}km = \sum_{i} L_i \times IP_{reach,i}$$

where $L_i$ is the length of reach $i$ in kilometers. IP-km describes a watershed in terms of its expected equivalent length of continuous high-quality habitat: a watershed with 30 IP-km does not necessarily have 30 km of uniformly excellent habitat, but the weighted integral of lower-quality habitat sums to what would be expected from 30 km of optimal habitat (Bjorkstedt et al. 2005).

> **Note:** All reaches are currently treated as accessible to anadromy. Once a barrier/passage layer is available, the IP-km sum should be restricted to accessible reaches.

```{r suitability-fns, include=FALSE}
suit_gradient <- function(s) {
  case_when(
    s <= 2          ~ 1.00,
    s > 2 & s <= 4  ~ 0.45,
    s > 4           ~ 0.00,
    TRUE            ~ NA_real_
  )
}

suit_discharge <- function(q) {
  case_when(
    q >= 3  ~ 1.00,
    q >= 1  ~ 0.45,
    q <  1  ~ 0.00,
    TRUE    ~ NA_real_
  )
}

suit_confinement <- function(vwi) {
  case_when(
    is.na(vwi) ~ NA_real_,
    vwi >= 4   ~ 1.00,
    vwi >= 1   ~ 0.45,
    vwi <  1   ~ 0.00
  )
}
```

```{r compute-ip, include=FALSE}
ip <- flowlines_ip |>
  mutate(
    reach_length_km = as.numeric(st_length(st_transform(geometry, 26910))) / 1000,
    suit_grad       = suit_gradient(slope),
    suit_q          = suit_discharge(discharge_mean_est),
    suit_conf       = suit_confinement(valley_width_index),
    IP_reach        = (suit_grad * suit_q * suit_conf)^(1/3)
  )

# Many reaches have IP_reach = 0 (geometric mean collapses when any suitability = 0),
# so Jenks on the full distribution produces duplicate breaks at 0.
# Instead: hard-classify zeros as Low, then apply Jenks to non-zero reaches only
# to find the Moderate / High / Very High boundaries.
breaks_jenks <- classIntervals(
  ip$IP_reach[!is.na(ip$IP_reach) & ip$IP_reach > 0],
  n = 3, style = "jenks"
)$brks

ip <- ip |>
  mutate(
    ip_class = case_when(
      is.na(IP_reach)             ~ NA_character_,
      IP_reach == 0               ~ "Low",
      IP_reach <= breaks_jenks[2] ~ "Moderate",
      IP_reach <= breaks_jenks[3] ~ "High",
      TRUE                        ~ "Very High"
    ),
    ip_class = factor(ip_class, levels = c("Very High", "High", "Moderate", "Low"))
  )
```

### Reach-scale Results

```{r ip-reach-table}
ip |>
  st_drop_geometry() |>
  filter(!is.na(IP_reach)) |>
  group_by(gnis_nm) |>
  summarise(
    Reaches  = n(),
    Min      = round(min(IP_reach),    2),
    Median   = round(median(IP_reach), 2),
    Mean     = round(mean(IP_reach),   2),
    Max      = round(max(IP_reach),    2),
    .groups  = "drop"
  ) |>
  arrange(desc(Mean)) |>
  knitr::kable(caption = "IP_reach distribution by stream (geometric mean of gradient, discharge, and valley confinement suitabilities).")
```

### Watershed-scale Results (IP-km)

```{r ip-km-table}
ip_km <- ip |>
  filter(!is.na(IP_reach)) |>
  st_drop_geometry() |>
  group_by(gnis_nm) |>
  summarise(
    `IP-km`              = round(sum(reach_length_km * IP_reach, na.rm = TRUE), 1),
    `Total length (km)`  = round(sum(reach_length_km, na.rm = TRUE), 1),
    `Mean IP_reach`      = round(mean(IP_reach, na.rm = TRUE), 2),
    Reaches              = n(),
    .groups              = "drop"
  ) |>
  arrange(desc(`IP-km`))

knitr::kable(
  ip_km,
  caption = "IP-km by stream: expected equivalent length of continuous high-quality habitat (Bjorkstedt et al. 2005)."
)
```

### Map

IP_reach classes are assigned using Jenks natural breaks on the observed 0–1 distribution.

```{r ip-map, fig.width=9, fig.height=7, fig.cap="Draft intrinsic potential for fall Chinook salmon spawning, Klamath Basin. Class breaks derived from Jenks natural breaks on IP_reach."}
ip_colors <- c(
  "Very High" = "#1a6b3c",
  "High"      = "#5aab61",
  "Moderate"  = "#f0c93a",
  "Low"       = "#d94f3d"
)

river_labels <- ip |>
  filter(!is.na(gnis_nm), gnis_nm != "") |>
  group_by(gnis_nm) |>
  summarise(geometry = st_union(geometry), .groups = "drop") |>
  st_point_on_surface() |>
  mutate(
    x = st_coordinates(geometry)[, 1],
    y = st_coordinates(geometry)[, 2]
  )

ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union(), fill = NA, color = "grey60") +
  geom_sf(data = ip, aes(color = ip_class), linewidth = 0.9) +
  scale_color_manual(values = ip_colors, name = "IP Class", na.translate = FALSE) +
  labs(
    title    = "DRAFT — Fall Chinook Intrinsic Potential",
    subtitle = expression(IP[reach] == (S[gradient] %.% S[discharge] %.% S[confinement])^{1/3}),
    caption  = paste0(
      "Jenks natural breaks  ·  ",
      "Thresholds not yet calibrated to local fish data"
    )
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid   = element_blank(),
    axis.title   = element_blank(),
    plot.caption = element_text(size = 9, color = "grey40")
  )
```

---

## Validation

### Spawning Gravel Surveys

RES conducted spawning gravel patch surveys within the former reservoir footprint and select tributaries. These data provide an initial qualitative check on whether reaches flagged as high IP_reach align with documented spawning habitat.

Source: [Spawning Habitat Patch data](https://klamath-data-management-platform-klamath.hub.arcgis.com/maps/92cad013db7c445f87194f09e8fcc71e)

```{r spawning-gravel}
spawning_gravel <- sf::read_sf(
  here::here("data-raw", "shapes", "Spawning_Habitat_Patch", "Spawning_Habitat_Patch.shp")
)

spawning_gravel |>
  sf::st_drop_geometry() |>
  janitor::clean_names() |>
  group_by(survey_reac) |>
  summarise(`Total patch area (sq ft)` = round(sum(patch_size_c, na.rm = TRUE), 0)) |>
  arrange(desc(`Total patch area (sq ft)`)) |>
  knitr::kable(caption = "Spawning gravel patch area by survey reach.")
```

---

## References

Bjorkstedt, E.P., L.W. Spence, J.C. Garza, C. Michel, D. Leonard, N. Bill, J. Quinn, and P. Sonntag. 2005. *An analysis of salmonid population viability in California.* Report to the California Department of Fish and Game, Sacramento, CA.

Burnett, K.M., G.H. Reeves, D.J. Miller, S. Clarke, K. Vance-Borland, and K. Christiansen. 2007. Distribution of salmon-habitat potential relative to landscape characteristics and implications for conservation. *Ecological Applications* 17(1):66–80.

Bieger, K., H. Kling, J. Schürz, and H. Fürst. 2015. Introduction to SWAT+, a completely restructured version of the Soil and Water Assessment Tool. *JAWRA* 51:816–828.

Busch, D.S., et al. 2013. Landscape-level model of potential Chinook salmon habitat. NOAA Northwest Fisheries Science Center.

Beechie, T.J., E. Buhle, M. Ruckelshaus, A. Fullerton, and L. Holsinger. 2006. Hydrologic regime and the conservation of salmon life history diversity. *Biological Conservation* 130(4):560–572.
