---
title: "Intrinsic Potential for Chinook Salmon in the Klamath Basin"
author: "Maddee Wiggins (FlowWest)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3      
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(sf)
library(rivermile)
library(terra)
library(stars)


```

## Objective

Calculate the intrinsic potential for Chinook salmon in the Klamath basin rivers and streams

## Aggregate and Process Data

### Flowlines

Source:

-   NHDplus geodatabase downloaded from: <https://www.sciencebase.gov/catalog/item/5d30c29ae4b01d82ce84aa5e>

-   NHDplus user guide: <https://pubs.usgs.gov/of/2019/1096/ofr20191096.pdf>

```{r include=FALSE}
flowlines <- rivermile::processed_nhd 
```

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines, color = "darkblue", alpha = 0.5) +
  theme_minimal()
```

```{r include=FALSE}
flowlines_ip_layers <- flowlines |> 
  select(gnis_nm, gnis_id, nhdplusid = nhd_pl_, slope, tot_da_sq_km = tt_d_s_) |> 
  mutate(slope = slope * 100)
```

### Slope

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers |> filter(slope < 20), # did this for the visusal since most of the values are below 20%
          aes(color = slope), size = 1) +
  theme_minimal()
```

```{r echo=FALSE}
flowlines_ip_layers |> group_by(gnis_nm) |> 
  filter(slope < 100) |> # removes two outliers
  summarise(min_slope = round(min(slope),1),
            max_slope = round(max(slope),1), 
            mean_slope = round(mean(slope),1), 
            median_slope = round(median(slope), 1)) |> 
  st_drop_geometry() |> 
  knitr::kable()

```

### Total Drainage Area (sqkm)

This value is part of the NHD network

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers, aes(color = tot_da_sq_km)) +
  theme_minimal()
```

### Estimated Discharge

There is likely a better way to do this with USGS gages however since the Chinook suitability is 1 for a discharge greater than zero, it doesn't seem worthwhile.

**double check source** source: <https://ui.adsabs.harvard.edu/abs/2018AGUFMEP21D2294W/abstract#>:\~:text=Geomorphologists%20often%20use%20drainage%20area%20as%20a,exponent%20decreases%20as%20flood%20recurrence%20interval%20increases.

```{r}
estimate_discharge <- function(area_km2) {
  discharge <- 0.028 * area_km2^0.89
  return(discharge)
}

flowlines_ip_layers <- flowlines_ip_layers |> 
  mutate(discharge_mean_est = estimate_discharge(tot_da_sq_km))

```

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers, aes(color = discharge_mean_est)) +
  theme_minimal()
```

```{r echo=FALSE}
flowlines_ip_layers |> group_by(gnis_nm) |> 
  summarise(min_discharge_mean_est = round(min(discharge_mean_est),1),
            max_discharge_mean_est = round(max(discharge_mean_est),1), 
            mean_discharge_mean_est = round(mean(discharge_mean_est),1), 
            median_discharge_mean_est = round(median(discharge_mean_est), 1)) |> 
  st_drop_geometry() |> 
  knitr::kable()

```

### Bankfull Width

Source: Bieger model $2.76 * DA^.399$

```{r}
flowlines_ip_layers <- flowlines_ip_layers |> 
  mutate(bf_width_m = 2.76 * tot_da_sq_km^0.399)
```

```{r}
ggplot() +
  geom_sf(data = rivermile::klamath_hucs |> st_union()) +
  geom_sf(data = flowlines_ip_layers, aes(color = bf_width_m)) +
  theme_minimal()
```

### Valley Bottom Width

Calculated using Valley Confinement methodology outlined in [South Fork Eel River Basin Geomorphic Classification](https://watermanagement.ucdavis.edu/download_file/view_inline/144) and implemented through [Habistat](https://github.com/FlowWest/habistat/blob/main/data-raw/scripts/terrain-attributes.md#valley-bottom-width-via-slope-cutoff-method). Code adapted to fit the needs of this project.

```{r}
valley_bottom <- function(catchment, flowline, dem) {
  dem |> 
    st_crop(catchment) |> 
    st_as_stars() |> 
    starsExtra::slope() |>
    mutate(valley_bottom = if_else(slope<units::set_units(atan(0.25)*180/pi,"degrees"), 1, NA)) |>
    select(valley_bottom) |>
    st_as_sf(merge=TRUE) |>
    st_filter(st_zm(flowline)) 
}

valley_bottom_by_comid <- function(x, dem, catchment, flowlines) {
  c <- catchments |> 
    mutate(nhdplusid = as.character(nhdplusid)) |> 
    filter(nhdplusid==x) |> 
    st_transform(st_crs(dem))
  f <- flowlines |> 
    mutate(nhdplusid = as.character(nhdplusid)) |> 
    filter(nhdplusid==x) |> 
    st_transform(st_crs(dem))
  valley_bottom(c, f, dem)
}

valley_width_naive <- function(geom_valley_bottom) {
  geom_valley_bottom |> 
    terra::vect() |> 
    terra::width()
}

perpendicular_transect <- function(pair, transect_length) {
  p1 <- st_coordinates(st_sfc(pair[[1]]))
  p2 <- st_coordinates(st_sfc(pair[[2]]))
  
  x1 <- p1[1, "X"]
  y1 <- p1[1, "Y"]
  x2 <- p2[1, "X"]
  y2 <- p2[1, "Y"]
  
  midpoint_x <- (x1 + x2) / 2
  midpoint_y <- (y1 + y2) / 2
  
  orig_bearing <- atan2(y2 - y1, x2 - x1)
  perp_bearing <- orig_bearing + pi / 2
  
  radius <- transect_length / 2
  
  perp_x1 <- midpoint_x + radius * cos(perp_bearing)
  perp_y1 <- midpoint_y + radius * sin(perp_bearing)
  perp_x2 <- midpoint_x - radius * cos(perp_bearing)
  perp_y2 <- midpoint_y - radius * sin(perp_bearing)
  
  st_linestring(matrix(
    c(perp_x1, perp_y1,
      perp_x2, perp_y2),
    ncol = 2,
    byrow = TRUE
  ))
}

perpendicular_transects <- function(ls, transect_length) {
  if (inherits(ls, "sf")) {
    crs_use <- st_crs(ls)
    ls <- st_geometry(ls)
  } else if (inherits(ls, "sfc")) {
    crs_use <- st_crs(ls)
  } else if (inherits(ls, "sfg")) {
    crs_use <- NA
    ls <- st_sfc(ls)
  } else {
    stop("ls must be sf, sfc, or sfg")
  }
  
  coords <- st_coordinates(st_zm(ls))[, 1:2, drop = FALSE]
  
  if (nrow(coords) < 2) {
    stop("Need at least 2 coordinates to build transects")
  }
  
  point_pairs <- lapply(seq_len(nrow(coords) - 1), function(i) {
    list(
      st_point(coords[i, ]),
      st_point(coords[i + 1, ])
    )
  })
  
  perp_lines <- lapply(point_pairs, function(pair) {
    perpendicular_transect(pair, transect_length)
  })
  
  st_sfc(perp_lines, crs = crs_use)
}

valley_width_via_transects <- function(geom_valley_bottom, flowline, max_width) {
  transects <- perpendicular_transects(flowline, transect_length = max_width)
  if("sfg" %in% class(geom_valley_bottom)) {
    geom_valley_bottom <- st_sfc(geom_valley_bottom, crs=st_crs(transects))
  }
  clipped <- st_intersection(transects, geom_valley_bottom)
  return(mean(st_length(clipped)))
}

```

```{r include=FALSE}
dem <- stars::read_stars("../../klamathAgData/data-raw/DEM/dem_z12_utm10.tif")
catchments <- rivermile::nhd_catchments |> 
  mutate(nhdplusid = as.character(nhdplusid)) |> 
  st_transform(st_crs(dem))
```

```{r}
catch <- catchments |> 
  filter(nhdplusid == '50000400209014') 
flow <- flowlines_ip_layers |> 
  mutate(nhdplusid = as.character(nhdplusid)) |> 
  filter(nhdplusid == '50000400209014') |> 
  st_transform(st_crs(dem))

valley_polygon <- valley_bottom(catch, flow, dem)
valley_width_via_transects(valley_bottom(catch, flow, dem), flow, max_width=1000)

ggplot() + 
  geom_sf(data=catch) +
  geom_sf(data=valley_polygon, fill="lightblue") +
  geom_sf(data=flow, color="blue") +
  geom_sf(data=perpendicular_transects(flow, transect_length=1000) |> st_intersection(valley_polygon), color="red") 

```

```{r eval=FALSE, include=FALSE}
dem_crs <- st_crs(dem)

safe_vb_width_simple <- purrr::possibly(
  function(x) {
    out <- valley_width_naive(x)
    if (length(out) == 0 || is.nan(out)) NA_real_ else as.numeric(out)
  },
  otherwise = NA_real_
)

safe_vb_width_transect <- purrr::possibly(
  function(a, b) {
    out <- valley_width_via_transects(a, b, 1000)
    if (length(out) == 0 || is.nan(out)) NA_real_ else as.numeric(out)
  },
  otherwise = NA_real_
)

vb_fun <- function(data) {
  data |> 
    mutate(nhdplusid = as.character(nhdplusid)) |> 
    #filter(nhdplusid %in% c("50000400015484")) |>
    mutate(flowline = map(
      geometry,
      ~ st_sfc(st_zm(.x), crs = st_crs(flowlines_ip_layers)) |>
        st_transform(st_crs(dem)))) |> 
    st_drop_geometry() |>
    mutate(valley_bottom_geoms = map(
      nhdplusid,
      possibly(function(x) {
        vb <- valley_bottom_by_comid(x, dem, flowlines = flowlines_ip_layers, catchment = catchments)
        geom <- st_union(st_geometry(vb))
        st_set_crs(geom, st_crs(dem))
      },
      otherwise = st_sfc(st_polygon(), crs = st_crs(dem))))) |>
    mutate(
      vb_width_simple = map_dbl(valley_bottom_geoms, safe_vb_width_simple),
      vb_width_transect = map2_dbl(valley_bottom_geoms, flowline, safe_vb_width_transect)
    ) |> 
    select(nhdplusid, flowline, valley_bottom_geoms, vb_width_simple, vb_width_transect) |>   glimpse()
}

vb_1_500 <- flowlines_ip_layers |>
  head(500) |> 
  vb_fun()

saveRDS(vb_1_500, here::here("data-raw","valley_bottom_width_1_500.rds"))

vb_501_1000 <- flowlines_ip_layers |>
  slice(501:1000) |> 
  vb_fun()

saveRDS(vb_501_1000, here::here("data-raw","valley_bottom_width_501_1000.rds"))

vb_1001_1500 <- flowlines_ip_layers |>
  slice(1001:1500) |> 
  vb_fun()

saveRDS(vb_1001_1500, here::here("data-raw","valley_bottom_width_1001_1500.rds"))

vb_1501_2000 <- flowlines_ip_layers |>
  slice(1501:2000) |> 
  vb_fun()

saveRDS(vb_1501_2000, here::here("data-raw","valley_bottom_width_1501_2000.rds"))

vb_2001_3000 <- flowlines_ip_layers |>
  slice(2001:3000) |> 
  vb_fun()

saveRDS(vb_2001_3000, here::here("data-raw","valley_bottom_width_2001_3000.rds"))

vb_3001_4000 <- flowlines_ip_layers |>
  slice(3001:4000) |> 
  vb_fun()

saveRDS(vb_3001_4000, here::here("data-raw","valley_bottom_width_3001_4000.rds"))

vb_4001_4236 <- flowlines_ip_layers |>
  slice(4001:4326) |> 
  vb_fun()

saveRDS(vb_4001_4236, here::here("data-raw","valley_bottom_width_4001_4236.rds"))

```

```{r include=FALSE}

vb <- bind_rows(read_rds(here::here("data-raw","valley_bottom_width_4001_4236.rds")), 
                read_rds(here::here("data-raw","valley_bottom_width_3001_4000.rds")),
                read_rds(here::here("data-raw","valley_bottom_width_2001_3000.rds")),
                read_rds(here::here("data-raw","valley_bottom_width_1501_2000.rds")),
                read_rds(here::here("data-raw","valley_bottom_width_1001_1500.rds")),
                read_rds(here::here("data-raw","valley_bottom_width_501_1000.rds")),
                read_rds(here::here("data-raw","valley_bottom_width_1_500.rds")))


```

```{r include=FALSE}
flowlines_ip_layers_new <- flowlines_ip_layers |> 
  mutate(nhdplusid = as.character(nhdplusid)) |> 
  left_join(vb |> 
              select(-flowline, -valley_bottom_geoms)) |> 
  glimpse()
```

### Valley Width Index

$V = (valley-floor-width)/(active-channel-width)$

```{r}
flowlines_ip_layers_all <- flowlines_ip_layers_new |> 
  mutate(valley_width_index = vb_width_transect/bf_width_m)
```

```{r}
ggplot() +
  geom_sf(data = flowlines_ip_layers_all, aes(color = valley_width_index), size = 2) + 
  theme_minimal()
```

```{r echo=FALSE}
flowlines_ip_layers_all |> group_by(gnis_nm) |> 
  summarise(min_valley_width_index = round(min(valley_width_index, na.rm = T),1),
            max_valley_width_index = round(max(valley_width_index, na.rm = T),1), 
            mean_valley_width_index = round(mean(valley_width_index, na.rm = T),1), 
            median_valley_width_index = round(median(valley_width_index, na.rm = T), 1)) |> 
  st_drop_geometry() |> 
  knitr::kable()

```

### Suitability Criteria

Source: <https://www.fs.usda.gov/pnw/pubs/journals/pnw_2005_agrawal001.pdf>

In addition to the above source, we looked into the following literature to see if the suitability significantly varied. Overall, this seems consistent for Chinook Salmon.

-   <https://www.fs.usda.gov/pnw/pubs/journals/pnw_2013_busch001.pdf>

-   <https://pubs.usgs.gov/publication/ofr20231077/>

```{r}
chinook_suit <- tibble(
  variable = c(
    rep("gradient_percent", 4),
    rep("valley_constraint", 4),
    rep("discharge_m3_s", 4)
  ),
  value = c(
    0,  2, 3.7, 8,
    1, 2.5, 4, 8,
    0, 3, 100, 100
  ),
  suitability = c(
    1.0, 0.45, 0.0, 0.0,
    0.8, 0.9, 1.0, 1.0,
    0.0, 1, 1, 1.0
  )
)

ggplot() + 
  geom_line(data = chinook_suit, aes(x = value, y = suitability)) +
  facet_wrap(~variable, scales = "free_x") +
  theme_minimal()

```

## Calculate Intrinsic Potential

```{r include = FALSE}

slope_suit <- chinook_suit |> filter(variable == "gradient_percent") |> arrange(value)
valley_suit <- chinook_suit |> filter(variable == "valley_constraint") |> arrange(value)
q_suit <- chinook_suit |> filter(variable == "discharge_m3_s") |> arrange(value)

lookup_step <- function(x, breaks, suits) {
  idx <- findInterval(x, breaks, rightmost.closed = TRUE)
  suits[pmax(idx, 1)]
}

slope_fun <- function(x) lookup_step(x, slope_suit$value, slope_suit$suitability)
valley_fun <- function(x) lookup_step(x, valley_suit$value, valley_suit$suitability)
q_fun <- function(x) lookup_step(x, q_suit$value, q_suit$suitability)

```

```{r}
ip <- flowlines_ip_layers_all |> 
  mutate(
    suit_slope = slope_fun(slope),
    suit_valley = valley_fun(valley_width_index),
    suit_q = q_fun(discharge_mean_est),
    chinook_ip = (suit_slope * suit_valley * suit_q)^(1/3)
  ) |> 
  mutate(
    chinook_ip_class = case_when(
      is.na(chinook_ip) ~ NA_character_,
      chinook_ip < 0.25 ~ "low",
      chinook_ip < 0.5  ~ "moderate",
      chinook_ip < 0.75 ~ "high",
      TRUE ~ "very high"
    )
  )

```

```{r echo=FALSE}
ggplot() + 
  geom_sf(data = ip, aes(color = chinook_ip_class), size = 2) +
  theme_minimal()

ggplot() + 
  geom_sf(data = ip, aes(color =chinook_ip ), size = 2) +
  theme_minimal()
```

```{r echo=FALSE}
ip |> 
  group_by(gnis_nm) |> 
  summarise(mean_ip = mean(chinook_ip, na.rm = T),
            min_ip = min(chinook_ip, na.rm = T),
            max_ip = max(chinook_ip, na.rm = T)) |> 
  st_drop_geometry() |> 
  knitr::kable()

```

## Data Questions

1.  How do we want the final dataset to be structured? Is it reasonable to summarize by reach?

2.  Do the values seem reasonable? It seems to be heavily driven by slope and some by valley constraint

3.  Are there any other datasets that would be beneficial to include?

```{r}
knitr::knit_exit()
```

```{r}
nhd_flowline <- readRDS(here::here("data-raw", "shapes", "nhd_flowline_hr.rds"))

flowlines <- nhd_flowline |>
  st_set_geometry("shape") |> 
  st_intersection(klamath_hucs |> st_union())

names(flowlines)[names(flowlines) == "shape"] <- "geometry"
st_geometry(flowlines) <- "geometry"
```

```{r}
flow_net <- flowlines 
seed_comid <- 50000400214915
seed_row <- flow_net |>
  st_drop_geometry() |>
  filter(comid == seed_comid)

edges <- flow_net |>
  st_drop_geometry() |>
  select(comid, from_node, to_node) |> 
  filter(!is.na(to_node))

g <- graph_from_data_frame(
  d = edges |> select(from_node, to_node),
  directed = TRUE
)

comp <- components(g, mode = "weak")

node_membership <- tibble(
  node = names(comp$membership),
  component_id = as.integer(comp$membership)
)

flow_comp <- flow_net |>
  left_join(
    node_membership |> 
      rename(from_node = node) |> 
      mutate(from_node = as.numeric(from_node)),
    by = "from_node"
  )

seed_component <- flow_comp |>
  st_drop_geometry() |>
  filter(comid == seed_comid) |>
  pull(component_id)

flow_clean <- flow_comp |>
  filter(component_id == seed_component)

flow_filter <- flow_clean |> 
  filter(stream_orde >= 5) |> 
  filter(tot_da_sq_km >= 1)

# flowlines_clean <- flow_clean |> 
#   select(comid, gnis_name, lengthkm, stream_leve, stream_orde, tot_da_sq_km, slope, geometry) |> 
#   filter(gnis_name %in% rivermile::all_klamath_rivers_line$river)
```
