---
title: "S3 Flow to Habitat Relationships by River Km"
output:
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
runtime: shiny
---
<style>
/* --- Navbar --- */
.navbar {
  background-color: #6c757d !important;
  border-color: #6c757d !important;
}
.navbar .navbar-brand,
.navbar-nav > li > a {
  color: #ffffff !important;
}

/* --- Make the whole dashboard feel less "zoomed out" --- */
/* Chrome/Edge/Safari */
html, body {
  zoom: 1.12;               /* try 1.08–1.20 depending on taste */
}

/* Firefox fallback (Firefox ignores zoom) */
@-moz-document url-prefix() {
  body {
    transform: scale(1.12);
    transform-origin: top left;
    width: 89.3%;           /* 100/1.12 to avoid horizontal overflow */
  }
}

/* --- Use space better (optional but helps perception) --- */
.container-fluid {
  padding-left: 12px;
  padding-right: 12px;
}
.section.level3 > h3 {
  margin-top: 6px;
  margin-bottom: 8px;
  font-weight: 600;
}
.html-widget {
  margin-top: 0 !important;
  margin-bottom: 0 !important;
}
</style>

```{r setup, include=FALSE}
library(flexdashboard)
library(sf)
library(leaflet)
library(tidyverse)
```

**Use application by clicking on river points to see how the WUAs change across the Klamath River**

```{r eval=FALSE, include=FALSE}
load("~/Documents/git/klamath/s3-model-code/model-files/export/S3/data/HABAREA.rda")
hab_area <- HABAREA |> 
  janitor::clean_names() |> 
  glimpse()

load("~/Documents/git/klamath/s3-model-code/model-files/export/S3/data/HABITAT.rda")
habitat <- HABITAT |> 
  janitor::clean_names() |> 
  mutate(river_km = round(rkm_up, 0)) |> 
  select(-rkm_up)
  glimpse()
  
hab_area_by_rm <- habitat |> left_join(hab_area)

klamath_river <- rivermile::klamath_river |> 
  glimpse()

klamath_river_units <- klamath_river |> 
  mutate(river_km = round(river_km, 0)) |> 
  filter(river_km %in% hab_area_by_rm$river_km) |> 
  select(geometry = x, river_km) |> 
  group_by(river_km) |>
  slice(1) |>       
  ungroup() |> 
  glimpse()

join <- hab_area_by_rm |> left_join(klamath_river_units)

saveRDS(join, "data-raw/habitat_river_mile_join.rds")

```


```{r}

join <- readRDS("habitat_river_mile_join.rds")

units_sf <- join |>
  select(unit_num, river_km,  geometry) |>
  distinct(unit_num, .keep_all = TRUE) |>
  st_as_sf() 

init_unit <- units_sf$unit_num[[1]]

plot_df <- join |>
  st_drop_geometry() |>
  select(unit_num, flow_cfs, spn_m2, fry_m2, juv_m2) |>
  group_by(unit_num, flow_cfs) |>
  summarise(
    spn_m2 = max(spn_m2, na.rm = TRUE),
    fry_m2 = max(fry_m2, na.rm = TRUE),
    juv_m2 = max(juv_m2, na.rm = TRUE),
    .groups = "drop"
  )
```

Column {data-width = 40}
-----------------------------------------------------------------------

```{r}
leafletOutput("unit_map", height = "650px")

output$unit_map <- renderLeaflet({
  leaflet(units_sf) |>
  addProviderTiles(providers$CartoDB.Positron, group = "Positron") |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Aerial Imagery") |>
  addLayersControl(
    baseGroups = c("Positron", "Aerial Imagery"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addCircleMarkers(
    radius = 5,
    stroke = TRUE,
    weight = 1,
    color = "green",      
    opacity = 0.9,
    fillColor = "darkgreen",
    fillOpacity = 0.55,
    layerId = ~unit_num,
    popup = ~paste0(
      "<div style='font-family: -apple-system, BlinkMacSystemFont, Segoe UI, Roboto, Helvetica, Arial, sans-serif;
      font-size: 13px; line-height: 1.3;'>",
      "<div style='margin-bottom: 6px;'>Unit Number: ", unit_num, "</div>",
      "<table style='border-collapse: collapse;'>",
      "<tr><td style='padding: 2px 10px 2px 0; color: #555;'>River Km:</td>",
      "<td style='padding: 2px 0;'>", river_km, "</td></tr>",
      "</table>",
      "</div>"
    ),
    popupOptions = popupOptions(
      closeButton = TRUE,
      maxWidth = 260
    )
  )

})

```

Column {data-width=60}
-----------------------------------------------------------------------
## Plot 

```{r}
selected_unit <- reactiveVal(init_unit)

observeEvent(input$unit_map_marker_click, {
  click <- input$unit_map_marker_click
  if (!is.null(click$id)) selected_unit(as.integer(click$id))
})

plotOutput("hab_plot", height = "650px")

output$hab_plot <- renderPlot({
  u <- selected_unit()

  df_u <- plot_df |>
    filter(unit_num == u) |>
    mutate(flow_cfs = as.numeric(flow_cfs)) |>
    pivot_longer(
      cols = c(spn_m2, fry_m2, juv_m2),
      names_to = "life_stage",
      values_to = "habitat_m2"
    ) |>
    mutate(
      life_stage = recode(
        life_stage,
        spn_m2 = "Spawn",
        fry_m2 = "Fry",
        juv_m2 = "Juvenile"
      )
    )

  ggplot(df_u, aes(x = flow_cfs, y = habitat_m2)) +
    geom_line(linewidth = 1) +
    facet_wrap(~life_stage, ncol = 1, scales = "free_y") +
    labs(
      title = paste("Unit", u),
      x = "Flow (cfs)",
      y = "Habitat (m²)"
    ) +
    theme_minimal() + 
    theme(
      strip.text = element_text(size = 16, face = "bold"),  
      axis.title = element_text(size = 15),
      axis.text = element_text(size = 13),
      plot.title = element_text(size = 16, face = "bold")
    )
})
```

